---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: yes
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "figure/")
```

## Loading and preprocessing the data

```{r}
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?

```{r}
steps_per_day <- tapply(activity$steps, activity$date, sum, na.rm = TRUE)
```

```{r}
hist(steps_per_day,
     main = "Total Steps per Day",
     xlab = "Steps",
     col = "steelblue")
```

```{r}
mean_steps <- mean(steps_per_day, na.rm = TRUE)
median_steps <- median(steps_per_day, na.rm = TRUE)
mean_steps
median_steps
```

## What is the average daily activity pattern?

```{r}
avg_interval <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
```

```{r}
plot(names(avg_interval), avg_interval,
     type = "l",
     xlab = "5-minute interval",
     ylab = "Average number of steps",
     main = "Average Daily Activity Pattern")
```

```{r}
max_interval <- names(avg_interval)[which.max(avg_interval)]
max_interval
```

## Imputing missing values

### Total missing values

```{r}
total_NA <- sum(is.na(activity$steps))
total_NA
```

### Explanation
The missing values in the dataset were imputed using the mean number of steps for each 5‑minute interval. This approach preserves the overall daily activity pattern because it replaces missing values with typical behavior for that specific interval rather than using a single global mean.

### Create imputed dataset

```{r}
interval_means <- avg_interval
activity_imputed <- activity

interval_df <- data.frame(
  interval = as.numeric(names(interval_means)),
  mean_steps = as.numeric(interval_means)
)

activity_imputed <- merge(activity_imputed, interval_df, by = "interval", all.x = TRUE)

na_idx <- is.na(activity_imputed$steps)
activity_imputed$steps[na_idx] <- activity_imputed$mean_steps[na_idx]

activity_imputed <- activity_imputed[
  order(activity_imputed$date, activity_imputed$interval),
  c("steps", "date", "interval")
]
```

### Histogram after imputation

```{r}
steps_per_day_imp <- tapply(activity_imputed$steps, activity_imputed$date, sum)

hist(steps_per_day_imp,
     main = "Total Steps per Day (Imputed)",
     xlab = "Steps",
     col = "darkgreen")
```

### Mean and median after imputation

```{r}
mean_steps_imp <- mean(steps_per_day_imp)
median_steps_imp <- median(steps_per_day_imp)
mean_steps_imp
median_steps_imp
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
day_names <- weekdays(activity_imputed$date)
day_type <- ifelse(day_names %in% c("Saturday", "Sunday"), "weekend", "weekday")
activity_imputed$day_type <- factor(day_type)
```

```{r}
avg_weekday <- tapply(activity_imputed$steps[activity_imputed$day_type == "weekday"],
                      activity_imputed$interval[activity_imputed$day_type == "weekday"],
                      mean)

avg_weekend <- tapply(activity_imputed$steps[activity_imputed$day_type == "weekend"],
                      activity_imputed$interval[activity_imputed$day_type == "weekend"],
                      mean)
```

```{r}
par(mfrow = c(2,1))

plot(names(avg_weekday), avg_weekday,
     type = "l",
     main = "Weekday Activity Pattern",
     xlab = "Interval",
     ylab = "Average Steps")

plot(names(avg_weekend), avg_weekend,
     type = "l",
     main = "Weekend Activity Pattern",
     xlab = "Interval",
     ylab = "Average Steps")
```

### Interpretation
Weekdays typically show a sharper morning peak, while weekend activity is more evenly distributed throughout the day.
